Supporting Information

### 2021-09-02 vertical-thermal-review.Rmd
minor fixes to legends

![**Figure `r fig_schematic`. Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature, (3) leaf traits, (4) leaf metabolism, and (5) whole-tree and ecosystem ecology. Abbreviations are as follows: VPD: vapor pressure deficit;  Tleaf - Tair: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; temp sensitivity: temperature sensitivity; ET: evapotranspiration, GPP: gross primary productivity.** Here, we summarize conditions in dense canopies during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open canopies, when canopy trees are seasonally deciduous, or at nighttime. Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27). ](schematics/Schematic figure 1.png)

@@

**Canopy trees account for the majority of forest ecosystem water and carbon cycling.**
While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011]. Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a; @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018]. In terms of C cycling, it has been estimated that canopy strata contribute $\geq$ 64% of net daytime ~CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007]. Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig. `r fig_ecosystem`, Piponiot et al., accepted).

![**Figure `r fig_ecosystem`.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests. Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 42 tropical forests from Piponiot et al. (accepted).](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)

@@

![**Figure `r fig_model`.** Lidar and thermal remote sensing data can be used to measure vertical canopy structure and plant thermal signatures at increasingly large scales. (a-b) Integrating in situ and remote sensing measurements will support our ability to scale understanding of vertical gradients within and across ecosystems. (c) In turn, these efforts can be used to directly inform ecosystem model development and parameterization. Vertically resolved in situ data could include any variable listed in Table 1 or 2. [Issue #42](https://github.com/EcoClimLab/vertical-thermal-review/issues/42). ](schematics/fig_remote_sensing.png)

>>>>

![**Figure `r fig_schematic`. Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature, (3) leaf traits, (4) leaf metabolism, and (5) whole-tree and ecosystem ecology. Abbreviations are as follows: VPD: vapor pressure deficit;  Tleaf - Tair: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; temp sensitivity: temperature sensitivity; ET: evapotranspiration, GPP: gross primary productivity.** Here, we summarize conditions in dense canopies during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open canopies, when canopy trees are seasonally deciduous, or at nighttime. Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy. ](schematics/Schematic figure 1.png)

@@

**Canopy trees account for the majority of forest ecosystem water and carbon cycling.**
While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011]. Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a; @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018]. In terms of C cycling, it has been estimated that canopy strata contribute $\geq$ 64% of net daytime ~CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007]. Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests [Fig. `r fig_ecosystem`b, @piponiot_sizerelated_inreview].

![**Figure `r fig_ecosystem`.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests. Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 42 tropical forests from @piponiot_sizerelated_inreview.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)

@@

![**Figure `r fig_model`.** Lidar and thermal remote sensing data can be used to measure vertical canopy structure and plant thermal signatures at increasingly large scales. (a-b) Integrating in situ and remote sensing measurements will support our ability to scale understanding of vertical gradients within and across ecosystems. (c) In turn, these efforts can be used to directly inform ecosystem model development and parameterization. Vertically resolved in situ data could include any variable listed in Table 1 or 2.](schematics/fig_remote_sensing.png)
